### 2025-07-19 paper/quarto_aaai/paper.tex
editorial pass until section 2

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals during the training
phase to minimize the divergence between learned representations and
plausible, actionable explanations. We demonstrate empirically and
theoretically that our proposed method facilitates training models that
deliver inherently desirable explanations while promoting robustness and
preserving high predictive performance of the models.
\end{abstract}
\section{Introduction}\label{introduction}

@@

Counterfactual explanations (CE) have become a key part of the broader
explainable AI (XAI) toolkit \citep{molnar2022interpretable} that can be
applied to make sense of this complexity. Originally proposed by
\citet{wachter2017counterfactual}, CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output (Figure~\ref{fig-poc}).
This is useful and necessary to not only understand how opaque models
make their predictions, but also to provide algorithmic recourse (AR) to
individuals subjected to them: a retail bank, for example, could use CE
to provide meaningful feedback to unsuccessful loan applicants that were
rejected based on an opaque automated decision-making (ADM) system.

For such feedback to be meaningful, counterfactual explanations need to
fulfill certain desiderata
\citep{verma2020counterfactual, karimi2020survey}---they should be
faithful to the model \citep{altmeyer2024faithful}, plausible
\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
Plausibility is typically understood as counterfactuals being
\emph{in-domain}: unsuccessful loan applicants that implement the
provided recourse should end up with credit profiles that are genuinely
similar to that of individuals who have successfully repaid their loans
in the past. Actionable explanations comply with practical constraints:
a young, unsuccessful loan applicant cannot increase their age in an
instant.

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because the narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore embrace the
paradigm that models---as opposed to explanation methods---should be
held accountable for explanations that are plausible and actionable.
While previous work has shown that at least plausibility can be
indirectly achieved through existing techniques aimed at models'
generative capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

@@

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) models' adversarial robustness improves
across the board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
classifiers trained under different regimes on synthetic data: (a)
conventional training, all mutable; (b) CT, all mutable; (c)
conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
linear decision boundary is shown in green along with training data
colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
(orange). Stars indicate counterfactuals in the target class. For
illustration, synthetic features are annotated as \emph{debt} and
\emph{age}.}

@@

To make the desiderata for our framework more concrete, we follow the
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to ``explanatory capacity'' as ``explainability'' in the rest
of this manuscript (see Def. \ref{def-explainability}).

>>>>

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
feature mutability constraints. Much existing research has therefore
focused on developing post-hoc methods to generate counterfactuals that
meet these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals during the training phase to minimize the divergence
between learned representations and plausible, actionable explanations.
We demonstrate empirically and theoretically that our proposed method
facilitates training models that deliver inherently desirable
explanations and have improved adversarial robustness.
\end{abstract}
\section{Introduction}\label{introduction}

@@

Counterfactual explanations (CE) have become a key part of the broader
explainable AI (XAI) toolkit \citep{molnar2022interpretable} that can be
applied to make sense of this complexity. Originally proposed by
\citet{wachter2017counterfactual}, CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output. This is useful and
necessary to not only understand how opaque models make their
predictions, but also to provide algorithmic recourse (AR) to
individuals subjected to them: a retail bank, for example, could use CE
to provide meaningful feedback to unsuccessful loan applicants that were
rejected based on an opaque automated decision-making (ADM) system
(Figure~\ref{fig-poc}).

For such feedback to be meaningful, counterfactual explanations need to
fulfill certain desiderata
\citep{verma2020counterfactual, karimi2020survey}---they should be
faithful to the model \citep{altmeyer2024faithful}, plausible
\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
Plausibility is typically understood as counterfactuals being
\emph{in-domain}: unsuccessful loan applicants that implement the
provided recourse should end up with credit profiles that are genuinely
similar to that of individuals who have successfully repaid their loans
in the past. Actionable explanations comply with practical constraints:
a young, unsuccessful loan applicant cannot increase their age in an
instance.

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because the narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore embrace the
paradigm that models (as opposed to explanation methods) should be held
accountable for explanations that are plausible and actionable. While
previous work has shown that at least plausibility can be indirectly
achieved through existing techniques aimed at models' generative
capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

@@

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to respect global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) models' adversarial robustness improves
across the board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
classifiers trained under different regimes on synthetic data: (a)
conventional training, all mutable; (b) CT, all mutable; (c)
conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
linear decision boundary is shown in green along with training data
colored according to ground-truth labels: \(y^-=\text{"loan withheld"}\)
(blue) and \(y^+=\text{"loan provided"}\) (orange). Class and feature
annotations (\emph{debt} and \emph{age}) are for illustrative purposes.}

@@

To make the desiderata for our framework more concrete, we follow
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to ``explanatory capacity'' as ``explainability'' in the rest
of this manuscript (see Def. \ref{def-explainability}).

### 2025-07-19 paper/sections/abstract.qmd
editorial pass until section 2

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the user-defined mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance of the models.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the feature mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations and have improved adversarial robustness.

### 2025-07-19 paper/sections/related.qmd
editorial pass until section 2

To make the desiderata for our framework more concrete, we follow the previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial]. For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def. \ref{def-explainability}).

>>>>

To make the desiderata for our framework more concrete, we follow previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial]. For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def. \ref{def-explainability}).
